|

### 2025-11-26 manuscript/pre_print.qmd
Edits to intro to better explain impact on recommendations

The consequences of widespread acceptance that REE is substantially lower in women living with PCOS should not be underestimated, particularly in light of the aforementioned incidence of overweight and obesity in this population. From a physiological perspective, if women with PCOS do exhibit a lower REE, this could imply a meaningful metabolic disadvantage that may influence dietary and nutritional guidance for weight management. However, if no such difference exists, the belief in a “slower metabolism” could instead serve as a deterrent to energy restriction based weight-loss approaches that are recognised as efficacious for improving PCOS symptoms [@aleneziImpactDietInducedWeight2024; @holteRestoredInsulinSensitivity1995] and are routinely recommended [@teedeRecommendations2023International2023]. It has been well documented that women with PCOS already experience higher rates of anxiety, depression, and lower quality of life (QOL) as a result of negative body image and weight-related concerns [@himeleinDepressionBodyImage2006; @hofmannBodyImageMental2025; @gellerBodyImageIllness2025; @davitadzeBodyImageConcerns2023]. Consequently, they are often observed to engage in more frequent weight-loss attempts than women without PCOS, though it was unclear whether this is focused on energy restriction approaches [@pesonenPolycysticOvarySyndrome2023]. Clarifying the relationship between REE and PCOS may therefore help guide more accurate clinical recommendations and empower both practitioners and women with PCOS.

@@

TO WRITE

>>>>

The consequences of widespread acceptance that REE is substantially lower in women living with PCOS should not be underestimated, particularly in light of the aforementioned incidence of overweight and obesity in this population. Women with PCOS typically engage in more frequent weight-loss attempts than women without PCOS [@pesonenPolycysticOvarySyndrome2023].  From a physiological perspective, if women with PCOS do exhibit a lower REE, this could imply a meaningful metabolic disadvantage that may influence dietary and nutritional guidance for weight management; for example, recommending a slightly more severe energy restriction to overcome the belief that they have a lower REE [@ozgensaydamWeightManagementStrategies2021]. Recommendations such as this could influence the well documented prevalence of eating disorders in women with PCOS [@lalonde-besterPrevalenceEtiologyEating2024; @jeanesBingeEatingBehaviours2017]. Contrastingly, belief in a “slower metabolism” could instead serve as a deterrent to energy restriction based weight-loss approaches for some women in line with typical general population recommendations that are also recognised as efficacious for improving PCOS symptoms [@aleneziImpactDietInducedWeight2024; @holteRestoredInsulinSensitivity1995] and are routinely recommended [@teedeRecommendations2023International2023]. It has been well documented that women with PCOS already experience higher rates of anxiety, depression, and lower quality of life (QOL) as a result of negative body image and weight-related concerns [@himeleinDepressionBodyImage2006; @hofmannBodyImageMental2025; @gellerBodyImageIllness2025; @davitadzeBodyImageConcerns2023]. Clarifying the relationship between REE and PCOS may therefore help guide more accurate clinical recommendations and empower both practitioners and women with PCOS.

@@

This study sought to estimate and describe the magnitude of difference in REE between women with and without PCOS. Most studies identified in the systematic review, and included in the meta-analysis, used indirect calorimetry as the primary measure of REE and assessed women with PCOS who were insulin resistant and categorised as being in overweight or obese BMI categories compared to healthy controls. Our results indicate there is only a small magnitude of difference in REE (`r round(main_arm_mean_effects_contrast_condition$draw)` kcal/day [95% quantile interval: `r round(main_arm_mean_effects_contrast_condition$.lower)` kcal/day to `r round(main_arm_mean_effects_contrast_condition$.upper)`kcal/day]) between women with PCOS and those without and also little difference in between person variation between the groups based on the ratio of standard deviations (`r round(main_arm_variance_effects_contrast_condition$draw,2)` [95% quantile interval: `r round(main_arm_variance_effects_contrast_condition$.lower,2)` to `r round(main_arm_variance_effects_contrast_condition$.upper,2)`]). These findings challenge the widely held belief that PCOS is inherently associated with a slower metabolism (PMID: 32244780), predisposing women with PCOS to weight gain. This belief may mistakenly lead to recommendations centred on slightly more severe calorie restriction to achieve weight-loss goals, compared with recommendations for the general population, as primary management strategies for women with PCOS [@ozgensaydamWeightManagementStrategies2021]. Recognising that there may be minimal differences in REE between women with PCOS and those without can inform both clinical and public practices, potentially leading to a shift in focus away from a requirement for more severe caloric restriction as a primary method of treatment towards more comprehensive, individualised, and psychologically safe approaches to care [@teedeRecommendations2023International2023].

The current study suggests that REE may not be the primary barrier to weight regulation in PCOS given small group-level differences between women with PCOS and those without (1442 kcal/day vs. 1472 kcal/ day, respectively). Similar to our findings, a case-control study comparing women with PCOS to healthy controls demonstrated REE was similar between groups (PMID: 38867472); however, REE adjusted for fat-free mass was lower in women with PCOS suggesting the importance of fat-free tissue in energy regulation. Indeed, a recent systematic review of mechanisms for metabolic dysfunction found excess androgen drives metabolic issues within adipose tissue and muscle tissue contributing to complications like obesity and insulin resistance (PMID: 32244780). Taken together, these findings highlight that factors beyond REE, such as hormonal and tissue-specific metabolic effects, likely play a more significant role in weight regulation challenges in women with PCOS.

Women with PCOS are routinely encouraged to participate in caloric restriction for weight management, advice that may inadvertently foster disordered eating. The prevalence of any eating disorder in women with PCOS has been reported at up to 62%, with affected women three to six times more likely than controls to experience an eating disorder (PMID: 38408541). Specific conditions such as binge-eating disorder (about 17%), bulimia nervosa (about 6 per cent) and night-eating syndrome (about 13%) are all markedly more common than in the general female population (PMID: 28104244, PMID: 27244869). Approximately 60% of women with PCOS report binge or comfort eating, with 39% displaying clinically significant behaviours (PMID: 27825940). 
The pathways linking PCOS and disordered eating are multifactorial. Biological mechanisms such as hyperandrogenism, hyperinsulinaemia, and altered ghrelin and leptin signalling can heighten hunger, carbohydrate cravings, and appetite variability (PMID: 21421016). Frequent hypoglycaemia and associated mood changes have also been observed, which can trigger compensatory eating or binge episodes (PMID: 21421016). These physiological processes interact with psychological and social stressors, including infertility concerns, conflicting nutrition advice, chronic dieting, and exposure to idealised body images on social media, which together compound vulnerability to disordered eating (PMID 33153014). Moreover, eating disorders themselves can disrupt endocrine function, potentially worsening PCOS symptoms and creating a self-reinforcing cycle (PMID 39115340, PMID 27244869). Understanding these intertwined biological, psychological, and social influences underscores the importance of avoiding overly restrictive dietary advice that can exacerbate feelings of failure, hunger dysregulation, and shame (PMID31615157). When metabolic rate is not impaired, recommending large caloric deficits is not only unnecessary but may be actively harmful.
A further barrier to effective management is weight stigma, which remains pervasive in both healthcare and everyday life (PMID: 39462252). Women with PCOS often report being labelled as “non-compliant” or “lacking willpower” when they do not lose weight rapidly, despite evidence of complex metabolic and psychosocial contributors (PMID: 37580314). Such attitudes are not benign. Weight stigma has been linked to poorer self-esteem, depression, anxiety, avoidance of medical care, and increased disordered eating. Physiologically, stigma-related stress elevates chronic cortisol, blood pressure, and insulin resistance, thereby worsening cardiometabolic outcomes (PMID: 37580314). Repeated cycles of weight loss and regain, driven by shame and unrealistic expectations, are associated with higher long-term weight, adverse lipid profiles, and elevated cardiovascular risk (PMID 37436934, PMID 23046078, PMID 10792339). Encouragingly, the 2023 international PCOS guidelines now recognise weight stigma as a determinant of health and call for its reduction across clinical and public health settings. Addressing stigma is therefore both an ethical imperative and a practical strategy to improve health outcomes.
Given these insights, the finding of a similar REE  in women with PCOS compared to those without PCOS invites a paradigm shift towards weight-neutral, individualised, and empowering care. Without evidence of metabolic disadvantage, interventions need not centre on extreme energy deficits. Instead, practitioners can focus on improving insulin sensitivity, inflammatory markers, and hormonal regulation through balanced nutrition, physical activity for enjoyment, adequate sleep, and stress management. Screening for disordered eating and providing evidence-based psychological support should be standard components of PCOS care. Approaches grounded in intuitive or mindful eating encourage self-trust and attunement to hunger and fullness cues, while dismantling moral judgements about food. This can stabilise glycaemic control, reduce binge episodes, and promote sustainable behaviour change (PMID: 32006391). Health coaching and dietetic practice should also include ongoing reflection on personal biases, the use of non-stigmatising language, and the creation of inclusive environments for people in all body sizes.
Importantly, this reframing is empowering for women with PCOS. Understanding that they are not “metabolically disadvantaged” enables them to move forward in their health journey without the burden of restrictive calorie targets or the belief that they are inherently at fault. By shifting attention towards metabolic-health parameters such as insulin resistance, stress management, and healthy lifestyle, women can pursue improvements in energy, mood, fertility, and quality of life on their own terms. When the focus becomes nourishment rather than restriction, self-care rather than self-control, and progress rather than perfection, sustainable wellbeing becomes achievable.

Strengths and Limitations
The current study has multiple strengths stemming from its preregistered, comprehensive methodology and Bayesian approach. This statistical framework allowed us to incorporate studies with and without control groups to better estimate mean REE in women with and without PCOS and to perform multiple sensitivity analyses that confirmed the stability of our findings. However, a limitation here is that variability in methods across studies, such as differences in PCOS diagnostic criteria and REE testing protocols, may have influenced results. Furthermore, some studies controlled for body weight or body composition when reporting REE values, while others did not. We accounted for this by estimating or converting reported data to obtain unadjusted REE values across all groups, thereby reducing this variability. Another limitation is that, due to fewer total control groups than PCOS groups, informed priors were required in several statistical models. Additionally, some studies reported data inconsistencies that could not be clarified (e.g., Georgopoulos et al.) and were accounted for in our analysis. Finally, most included studies were cross-sectional or baseline assessments within intervention trials, which limits causal inference.

Conclusion
In conclusion, the findings from this meta-analysis indicate that resting energy expenditure does not meaningfully differ between women with and without PCOS. Group-level differences in resting energy expenditure were small, insignificant, or not physiologically relevant. Additionally, variability in resting energy expenditure was also similar, suggesting that a lower baseline metabolic rate is not a causal factor for the weight-related challenges often associated with PCOS. These findings challenge the popular narrative that women with PCOS have a lower resting metabolism and may help better inform dietary interventions and nutritional support for these individuals. Future research should include more standardized REE measurement and reporting protocols, greater data transparency, consistent control and reporting of body weight or body composition, the presentation of both absolute and relative REE, and more precise characterization of PCOS phenotypes. Overall, these findings support the conclusion that PCOS does not negatively influence resting energy expenditure and may help practitioners and researchers focus on targeted lifestyle interventions rather than negatively framed metabolic assumptions.
